\chapter*{Introduction}
% \label{chp:intro}
\addcontentsline{toc}{chapter}{Introduction}

\vcsn is a free software platform dedicated to the manipulation of
finite state automata.  Here, `finite state automata' is to be
understood in the broadest sense: \vcsn supports \emph{weighted}
automata over a free monoid, and even \emph{weighted} automata on some
\emph{non-free monoids} (currently only automata on products of two
free monoids--- also known as \emph{transducers}---are supported).


\bigskip

From user's point of view, the platform consists in two main components:

\begin{description}
\item[The {\textnormal{\textsc{Vaucanson}}} library] is a \Cpp library that implements objects for
  automata, rational expressions, as well as algorithms on these
  objects.  
  This library is written under the static genericity paradigm.

\item[\tafkit] is a \emph{command-line interface} to the library that allows
  user to execute \vcsn's algorithms without any knowledge of \Cpp 
  nor of \vcsn API.
  This  interface is instantiated for a
  predefined set of commonly used automaton types.
  
\end{description}

The platform is coupled with two other communication modules:

\begin{description}
\item[An \xml format for automata and expressions\textnormal{,}] called \fsmxml,
aims at being a general purpose interchange format for weighted
automata and regular expressions. 
It is used as the normal, and default, input and 
output format for \tafkit and thus for the communication between 
\tafkit and \vgi.

\item[A Graphical User Interface] called \vgi and especially dedicated to
\vcsn,
allows to describe automata and to visualize
the result of operations on automata in an interactive and graphical 
way.   
All functions
defined in \tafkit can be called via the menu of \vgi.

\end{description}

The \vcsn platform is currently under a thorough revision and its
core is rebuilt with a new design that changes the interface and the
associated API. The ultimate version of the first phase of the
project, coined \vcsnvo, has been 
eventually released in September~2011. 
The novelty of the \vcsnv version, just released in July~2012, is
the integration of the Graphical User Interface \vgi.


\medskip

Ideally, a user's manual for \vcsn should document all of these 
components. 
We decided not to do so, not so much because it is a lot of work, but 
also as this work would not be so useful since the new design is 
likely to make most of the description soon obsolete.
On the other hand, we want to have a version of the platform that 
will serve as  
a landmark for both functionalities and  
performance of the first phase of \vcsn.


This version is coined \vcsnv and we consider it is accessible 
through the \tafkit only.
There will be a \tafkit for the future versions of \vcsn,
its  functionalities will include all those of the present one,
and its interface will essentially be the same as the \tafkit of 
\vcsnv.
\tafkit will be the only documented part of \vcsnv.


% It is accessible through the
% \tafkit only, which is also the only part that is documented.
% presented at FSMNLP~2011, and

\tafkit stands for \emph{Typed Automata Function Kit}. In the static
generic programming paradigm used in the \vcsn library, the
\emph{types} of the automata that are treated have to be known at
compile time. The command-line interface \tafkit is a set of
programs called from the \code{shell} and used to chain operations
on automata. At the installation of \vcsn, \tafkit is therefore
compiled for several predefined types of automata.

\tafkit does not allow to write new algorithms nor to manipulate new
types of automata, but it makes it possible, without efforts, to use
\emph{already programmed} functions on automata of \emph{predefined
types}. \tafkit gives a \emph{restricted access} to \vcsn
functionalities, but it is a \emph{direct access}, without any need
of programming skill. A basic familiarity with Unix command syntax
only is sufficient to make use of \tafkit.


The type of an automaton is determined by: (a) the fact it is an
automaton over a free monoid, or over a product of two free monoids,
(b) the type of the generators of the free monoid(s), (c) the weight
semiring. The weight semiring can be~$\B$, $\Z$, $\Q$, $\R$,
$\F_{2}$, $\Zmin$,   or~$\Zmax$. The generators may be
\emph{characters}, \emph{integers} or \emph{pairs} of these. At the
installation of \vcsn, \tafkit is instanciated for 18 combinations
of weights and generators types.

Besides basic editing commands, most of `classical' operations on
automata, together with less classical ones, are available in the
\tafkit instances: from transformations of automata into expressions
and back to operations such as quotient, product, and shuffle, from
operations for Boolean automata such as determinisation or
computation of the universal automaton to operations on transducers
such as composition or evaluation, from reduction of automata with
weights in a field to transformation of automata over pairs into
transducers.




This version \tafkitv has been presented at the CIAA~2012 conference held in 
Porto (Portugal).
All users are encouraged to send us remarks, comments, and bug reports.
We shall make our possible to take them into account in the minor 
revisions that will be made to \vcsnv until the release of \vcsn~2.0.

\endinput 





















% \bigskip
% 
% The platform consists in a few components:
% 
% \begin{description}
% \item[The \vcsn library] is a \Cpp library that implements objects for
%   automata, rational expressions, as well as algorithms on these
%   objects.  This library is generic, in the sense that it makes it
%   possible to write an algorithm once and apply it to different types
%   of automata.  However this genericity is achieved in a way that
%   should not cause any slowdown at runtime: because the type of the
%   automaton manipulated is known at compile time, compiling an
%   algorithm will generate code that is almost as efficient as an
%   algorithm written specifically for this type of automaton.
% 
% \item[\tafkit] is a command-line interface to the library that allows
%   user to execute \vcsn's algorithms without any knowledge of \Cpp.
%   Because the \vcsn library needs to know the type of automata at
%   compile time, the \tafkit interface has been instantiated for a
%   predefined set of common automaton types.
% 
%   \tafkit does not allow to write new algorithms nor to manipulate new
%   types of automata, but it makes it possible to combine without
%   efforts a large set of algorithms on common automata types.
% 
% \item[A repository of automata] that shows examples of automata of
%   various types, and also contains programs, called 
%   \emph{automata-factory}, which create parametrized families of
%   automata.
% 
% \end{description}
% It is coupled with some other modules:
% 
% \begin{description}
% \item[An \xml format for automata] and expressions, called \fsmxml.
% This format aims at being an interchange format for automata and thus 
% at making possible, and hopefully easy, the communication between 
% various programs that input or output automata.
% So far,  this format is used as the normal, and default, input and 
% output format for \tafkit.
% 
% \item[A graphic user interface] called \vgi, especially dedicated to
% \vcsn is under development at the EE Department of the National Taiwan
% University in Taipeh.  It will allow to describe automata and to visualize
% the result of operation on automata in a graphical way.  All functions
% defined in \tafkit will be called via the menu of \vgi.


% \end{description}
% 
% Ideally, a user's manual for \vcsn should document all of these components.
% We decided not to do so, not so much because it is a lot of work, but 
% also as this work would not be so useful.
% 
% After several years of hard and complex developments, the evolution 
% and progress of the \vcsn platform are now stuck and we have reached 
% the conclusion that we have to undertake a thorough revision of the 
% \vcsn library that will most probably change its interface and the one 
% of the associated API.
% These new developments will give rise to a new series of 
% versions of \vcsn, coined \vcsn 2.x.

% On the other hand, there will be a \tafkit for these future versions of 
% \vcsn, whose functionalities will include all those of the present 
% one and whose interface will essentially be the same as the present 
% one as well.
% \tafkit \vcsnv will serve as a landmark for both functionalities and 
% performance of the first version of \vcsn.
% It will be the only documented part of \vcsnv.


However, we want to have a version of the platform that will serve as 
a landmark for both functionalities and  
performance of the first phase of \vcsn.
It will be coined \vcsnv.
Moreover, there will be a \tafkit for the future versions of \vcsn,
its  functionalities will include all those of the present one,
and its interface will essentially be the same as the \tafkit of 
\vcsnv.
\tafkitv will be the only documented part of \vcsnv.

A beta version of \vcsnv has been presented at the FSMNLP 2011 Conference, 
held in Blois, France, from July 12 to July 15 2011.
All users are encouraged to send us remarks, comments, and bug reports.
We shall make our possible to take them into account in the minor 
revisions that will be made to \vcsnv until the release of \vcsn~2.0.

\endinput 


