\addcontentsline{toc}{chapter}{Bibilography}

\begin{thebibliography}{1}
    
	
\bibitem{AngrEtAl10}
{\sc P.-Y.~Angrand, S.~Lombardy, J.~Sakarovitch. }
\newblock On the number of broken derived terms of a rational 
expression, 
\newblock {\em J. Automata, Languages and Combinatorics\/}~{\bf 15} 
(2010), 27--51. 
		
\bibitem{AllaEtAl03}
{\sc C. Allauzen, M. Mohri, F. Pereira and M. D. Riley}
\newblock FSM Library Man pages.
\newblock {\tt http://www2.research.att.com/~fsmtools/fsm/} 
 AT\&T (1998--2003).

\bibitem{AllaEtAl10}
{\sc C. Allauzen and M. D. Riley}
\newblock OpenFst Library Wiki.
\newblock {\tt http://www.openfst.org/} (2010).

\bibitem{Anti96}
{\sc V. Antimirov. }
\newblock Partial derivatives of regular expressions and finite 
automaton constructions.
\newblock {\em Theoret. Comput. Sci.\/}~{\bf 155} (1996), 291--319. 

\bibitem{ClavEtAl05}
{\sc T.~Claveirole, S.~Lombardy, S.~O'Connor, L.-N.~Pouchet, and J.~Sakarovitch. }
\newblock Inside {V}aucanson.
\newblock {\em Proc. CIAA~2005}, LNCS~3845, Springer (2005) 117--128.

\bibitem{ChamZiad02}
{\sc J.-M. Champarnaud, D. Ziadi. }
\newblock Canonical derivatives, partial derivatives and finite 
automaton constructions.
\newblock {\em Theoret. Comput. Sci.\/}~{\bf 289} (2002) 137--163. 

\bibitem{ChenEtAl58}
{\sc K.~T. Chen, R.~H. Fox and R.~Lyndon. }
\newblock Free differential calculus {IV} -- {T}he
  quotient groups of the lower central series.
\newblock {\em Ann. Math.\/}~{\bf 68} (1958), 81--95.

\bibitem{Conw71}
{\sc J. H. Conway. } 
\newblock {\em Regular Algebra and Finite Machines}, 
\newblock Chapman and Hall, 1971. 

\bibitem{DelgMora04}
{\sc M. Delgado, J. Morais. }
\newblock  Approximation to the smallest regular 
expression for a given regular language. 
\newblock {\em Proc. CIAA 2004}, LNCS 3317, Springer (2004) 312--314. 

\bibitem{DemaEtAl08}
{\sc A. {Demaille}, A. {Duret-Lutz}, F. {Lesaint}, S. Lombardy,
J. Sakarovitch. and F. {Terrones}}
\newblock An XML format proposal for the description of weighted automata,
transducers, and regular expressions.
\newblock {\em Post-Proc. FSMNLP 2008} 
(J. Pikorski, B. Watson, A. Yli-Jyr\"a, eds.) 
IOS Press (2009) 199--206.

\bibitem{FrouSaka10}
{\sc Ch.~Frougny, J.~Sakarovitch. }
\newblock Number representation and finite automata,
\newblock in {\em Combinatorics, Automata, and Number Theory}
           (V.~Berthé, M.~Rigo , eds)
\newblock Cambridge University Press (2010) 34--107. 

\bibitem{LombSaka05a}
{\sc S.~Lombardy, J.~Sakarovitch. }
\newblock Derivatives of rational expressions with
multiplicity,
\newblock {\em Theoret. Comput. Sci.\/}~{\bf 332} (2005), 141--177. 

\bibitem{LombSaka07}
{\sc S.~Lombardy, J.~Sakarovitch. }
\newblock The Universal Automaton,
\newblock in {\em Logic and Automata, History and Perspectives}
(J.~Flum, E.~Gr\"adel, Th.~Wilke, eds)
\newblock Amsterdam Univ. Press, 2007, 457--504. 

\bibitem{LombSaka12}
{\sc S.~Lombardy, J.~Sakarovitch. }
\newblock The Removal of Weighted $\varepsilon$-Transitions,
\newblock {\em Proc. CIAA~2012} LNCS~7381, Springer (2012) 345--352.

\bibitem{LombSakaXX}
{\sc S.~Lombardy, J.~Sakarovitch. }
\newblock On the weighted closure problem,
\newblock {\em in preparation}.

\bibitem{Loth83}
{\sc M.~Lothaire. } 
\newblock {\em Combinatorics on {W}ords}, Addison-Wesley, 1983.
\newblock Reprint: Cambridge University Press, 1997.

\bibitem{Saka03}
{\sc J. Sakarovitch}
\newblock {\em \'El\'ements de th\'eorie des automates}.
\newblock Vuibert, 2003.
\newblock English corrected edition: \emph{Elements of Automata  
Theory}, Cambridge  University Press, 2009.

\bibitem{Saka09} 
{\sc J. Sakarovitch}
\newblock Rational and recognisable power series, 
\newblock In \emph{Handbook of Weighted Automata}
(M.~Droste, W.~Kuich and H.~Vogler, eds.) 
Springer (2009) 105--174.

\end{thebibliography}
\clearpage

\addcontentsline{toc}{chapter}{Index}
\printindex 
