
\chapter{Specification of functions on automata and rational 
expressions}
\label{chp:spe-fun}%

Functions are classified according to the type of automata they are 
applied to. 
They depend upon the type of the monoid: free monoid or direct 
product of two free monoids at this stage for \tafkitv and 
upon the type of the multiplicity semiring: `numerical' semirings
of different kinds.
% in 
% general, and $\R$ (implemented as \code{float}) as an example of a 
% \emph{field} and Boolean semiring in particular.
Some functions are specialised to even more particular type of 
alphabets.
Note that \tafkitv  offers no instance where the multiplicity 
semiring is a semiring of series (over 
a free monoid with multiplicity in a numerical semiring).\footnote{%
   Whereas this possibility is offered within the \emph{library} 
   \vcsnv, but it is out of the scope of this user's manual.}
In this chapter, we give the specifications of the functions, that 
is, the preconditions on their arguments, and the description of the 
result and how it is related to the argument.
% How this result is computed, if it does not influence the result, 
% will be described at \appnd{alg-spe}.

\begin{enumerate}
\item General automata and rational expressions 
\item Weighted  automata and rational expressions over free monoids
\item Automata and rational expressions over free monoids with 
weights in a field.
\item Boolean automata and rational expressions over free monoids
\item Weighted automata over product of two free monoids
\item Weighted automata over free monoids over alphabets of pairs
\end{enumerate}

This classification is used to organise the lists of commands.
Every instance of \tafkit contains the commands of the first section 
and of one or several others, as indicated in \tabla{tfk-ins-com} 
below.
A command with input and output arguments with different types 
belongs to the instance corresponding to the \emph{input} type.
Moreover, such a command exists only if the type of the \emph{output} 
argument is instanciated as well (\cf \Fct{partial-identity},  
\sbsct{par-ide}).

Every section begins with the list of commands that are then 
described in the section. 

\longonly{%
\begin{ComVd}{110725}
	This list is supplemented by the one of those commands which 
	finally have not been implemented in \tafkitv, or wich have been 
	implemented for a restricted class of automata, for the sake of 
	quotation in view of forthcoming versions. 
\end{ComVd}
}%

\begin{table}[ht]
\begin{center}
\begin{tabular}{ll>\e l>\ee l}
\hline
\multicolumn{1}{c}{command name} & 
\multicolumn{1}{c}{alphabet type} & 
\multicolumn{1}{c}{weight semiring} & 
\multicolumn{1}{c}{function sections} \\
\hline
\command{vcsn-char-b} 
% & automata 
& characters & $\langle\B,\lor,\land\rangle$ 
& 1, 2, 4  \\
\command{vcsn-int-b} 
% & automata 
& integers & $\langle\B,\lor,\land\rangle$ 
& 1, 2, 4  \\
\hline 
\command{vcsn-char-z} 
% & automata 
& characters & $\langle\Z,+,\times\rangle$
& 1, 2  \\
\command{vcsn-int-z} 
% & automata 
& integers & $\langle\Z,+,\times\rangle$
& 1, 2 \\
\command{vcsn-char-zmax} 
% & automata 
& characters & $\langle\Z,\max,+\rangle$
& 1, 2 \\
\command{vcsn-char-zmin} 
% & automata 
& characters & $\langle\Z,\min,+\rangle$
& 1, 2 \\
\command{vcsn-char-r} 
% & automata 
& characters & $\langle\R,+,\times\rangle$
& 1, 2, 3 \\
\command{vcsn-char-q} 
% & automata 
& characters & $\langle\Q,+,\times\rangle$
& 1, 2, 3 \\
\command{vcsn-char-f2} 
% & automata 
&  characters & $\langle\F_{2},+,\times\rangle$
& 1, 2, 3 \\
\hline 
\command{vcsn-char-char-b}\e 
% & automata 
& pairs of characters & $\langle\B,\lor,\land\rangle$
& 1, 2, 4, 6 \\
\command{vcsn-char-int-b} 
% & automata 
& pairs of character and integer & $\langle\B,\lor,\land\rangle$
& 1, 2, 4, 6 \\
\command{vcsn-int-int-b} 
% & automata 
& pairs of integers & $\langle\B,\lor,\land\rangle$
& 1, 2, 4, 6 \\
\command{vcsn-char-char-z} 
% & automata 
& pairs of characters & $\langle\Z,+,\times\rangle$
& 1, 2, 6 \\
\command{vcsn-int-int-z} 
% & automata 
& pairs of integers & $\langle\Z,+,\times\rangle$
& 1, 2, 6 \\
\hline
\command{vcsn-char-fmp-b} 
% & transducers 
& characters & $\langle\B,\lor,\land\rangle$
& 1, 5 \\
\command{vcsn-char-fmp-z} 
% & transducers 
& characters & $\langle\Z,+,\times\rangle$
& 1, 5 \\
\command{vcsn-int-fmp-b} 
% & transducers 
& integers & $\langle\B,\lor,\land\rangle$
& 1, 5 \\
\command{vcsn-int-fmp-z} 
% & transducers 
& integers & $\langle\Z,+,\times\rangle$
& 1, 5 \\
\hline
\end{tabular}
\end{center}
\caption{The \tafkit instances in \vcsnv and their commands}
\label{tab:tfk-ins-com}
\end{table}
\clearpage 


\input{TFKD-ch2-1}
\clearpage 
\input{TFKD-ch2-2}
\clearpage 
\input{TFKD-ch2-3}
\clearpage 
\input{TFKD-ch2-4}
\clearpage 
\input{TFKD-ch2-5}
\clearpage 
\input{TFKD-ch2-6}

\endinput
